Dependent Types for Safe and Secure Web Programming

### 2013-11-11 paper/introduction.tex

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
%\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage[numbers]{natbib}
%\WarningsOff

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\newcommand{\idata}{\textsf{iData}}
\newcommand{\itasks}{\textsf{iTasks}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{IFL '13}{August 28--31, 2013, Nijmegen, The Netherlands} 
\copyrightyear{2013} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

\title{Practical Dependently-Typed Programming for the Web}
\subtitle{Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages}
%\subtitle{Using Embedded Domain-Specific Languages to Improve Confidence in Web Applications}

\authorinfo{Simon Fowler \and Edwin Brady}
           {School of Computer Science, University of St Andrews, St Andrews, Scotland}
           {Email: \{sf37, ecb10\}@st-andrews.ac.uk}

\maketitle

\begin{abstract}
Dependently-typed languages allow very expressive types to be used during development, in turn facilitating easier reasoning about the operation of programs written in such languages. Stronger type specifications do however bring with them the disadvantage
that it becomes increasingly difficult to write programs that are accepted by
the type checker and additional proofs may have to be specified by a user.

Embedded domain-specific languages (EDSLs) address this problem by introducing
a layer of abstraction over more specific underlying types, allowing
domain-specific code to be written in high-level languages which use dependent
types to enforce certain invariants without additional proof obligations.

In this paper, we apply this technique to web programming, and introduce an
EDSL to facilitate the creation and handling of web forms which retain type
information, reducing the scope for programmer error and attacks such as SQL
injection. We also show how to enforce resource usage
protocols associated with common web operations such as CGI, database access
and session handling.

\end{abstract}

%\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

%\keywords
%keyword1, keyword2

@@

%
\begin{Verbatim}
  $conn = new mysqli("localhost", "username", 
                     "password", "db");
\end{Verbatim}
%
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it. However, this would cause
problems with later steps.

%
\begin{Verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{Verbatim}
%
We then create a prepared statement detailing our query, and bind the
\texttt{`dept'} value:

\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. 
Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.

%
\begin{Verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{Verbatim}
%
Finally, once the statement and connection are no longer needed, they should be
closed in order to discard the associated resources:

%
\begin{Verbatim}
  $stmt->close();
  $conn->close();
\end{Verbatim}
%
Even in this small example, there exists a specific resource usage protocol.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

@@

\idris{} \cite{brady2011idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework, which allows the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

@@

The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/SimonJF/IdrisWeb}.
% =================================================
% =================================================

\section{An overview of the \texttt{Effects} framework}
\label{effects}

\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
side-effects such as state, exceptions, and I/O as \emph{algebraic
effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
by an input and output \emph{state}, which permits effectful programs to track
the progress of a resource usage protocol. Effectful programs are written
in a monadic style, with \texttt{do}-notation, with their type stating which
specific effects are available.
Effectful
programs are described using the following data type,
in the simplest case:

\begin{Verbatim}
Eff : (m : Type -> Type) ->
      (es : List EFFECT) -> (a : Type) -> Type
\end{Verbatim}

\texttt{Eff} is parameterised over a \emph{computation context}, \texttt{m}, 
which describes the context in which the effectful program will be run, a
list of side effects \texttt{es} that the program is permitted to use, and the
program’s return type \texttt{a}. The name \texttt{m} for the computation context is
suggestive of a monad, but there is no requirement for it to be so.

For example, the following type carries an integer state,
throws an exception of type \texttt{String} if the state reaches 100, 
and runs in a \texttt{Maybe} context:

\begin{Verbatim}
addState : Eff Maybe [STATE Int, EXCEPTION String] ()
addState = do val <- get
              when (val == 100) (raise "State too big")
              put (val + 1)
\end{Verbatim}

\subsection{Implementing Effects}

Effects such as state and exception are described as algebraic data types,
and run by giving \emph{handlers} for specific computation contexts. 
Effects have a corresponding \emph{resource} (in the case of state, the
resource is simply the current state). Executing an effectful operation may
change the resource and return a value:

\begin{Verbatim}
Effect : Type
Effect = (in_res : Type) -> (out_res : Type) -> 
         (val : Type) -> Type
\end{Verbatim}

For example, the state effect is described as follows:

\begin{Verbatim}
data State : Effect where
     Get :      State a a a
     Put : b -> State a b ()
\end{Verbatim}

That is, \texttt{Get} returns a value of type \texttt{a} without updating
the resource type. \texttt{Put} returns nothing, but has the effect of updating
the resource. In order to make an effect usable, we implement a handler
for a computation context by making an instance of the following class:

\begin{Verbatim}
class Handler (e : Effect) (m : Type -> Type) where
     handle : res -> 
              (eff : e res res' t) -> 
              (k: res' -> t -> m a) -> m a
\end{Verbatim}

The \texttt{handle} function takes the input resource, an effect which may
update that resource and execute a side-effect, and a continuation \texttt{k}
which takes the updated resource and the return value of the effect. We use
a continuation here primarily because there is no restriction on the number of
times a handler may invoke the continuation (raising an exception, for example,
will not call the continuation). Reading and updating states in handled
for all computation contexts \texttt{m} as follows:

\begin{Verbatim}
instance Handler State m where
     handle st Get     k = k st st
     handle st (Put n) k = k n ()
\end{Verbatim}

Finally, we promote \texttt{State} into a concrete effect \texttt{STATE}, and
the \texttt{Get} and \texttt{Put} operations into functions in \texttt{Eff}, as
follows:

\begin{Verbatim}
STATE : Type -> EFFECT
STATE t = MkEff t State

get : Eff m [STATE x] x
get = Get

put : x -> Eff m [STATE x] ()
put val = Put val
\end{Verbatim}

A concrete effect is simply an algebraic effect type paired with its current
resource type. This, and other
technical details, are explained in full elsewhere~\cite{brady:effects}.
For the purposes of this paper, it suffices to know how to describe and
handle new algebraic effects.

\subsection{Resource Protocols as Effects}

More generally, a program might modify the set of effects available.
This might be desirable for several reasons, such as adding a new
effect, or to update an index of a dependently typed state. In this
case, we describe programs using the \texttt{EffM} data type:

\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) -> (es' : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}

\texttt{EffM} is parameterised over the context and type as before, but
separates input effects (\texttt{es}) from output effects (\texttt{es'}). 
In fact, \texttt{Eff}
is defined in terms of \texttt{EffM}, with equal input/output effects.
We can use this to describe how effects follow resource protocols. A simple
example is a file access protocol, where a file must be opened before it
is read or written, and a file must be closed on exit:

\begin{Verbatim}
data FileIO : Effect where
     Open  : String -> (m : Mode) -> 
             FileIO () (OpenFile m) ()
     Close : FileIO (OpenFile m) () ()

     ReadLine  : FileIO (OpenFile Read)  
                        (OpenFile Read) String
     WriteLine : String -> 
                 FileIO (OpenFile Write) 
                        (OpenFile Write) ()
     EOF       : FileIO (OpenFile Read)  
                        (OpenFile Read) Bool
\end{Verbatim}

Note that the types of the input and output resources describes how resource
state changes in each operation: opening a file changes an empty resource to
a resource containing an open file; reading a line is only possible if the
resoource is a file open for reading, etc.
The handler for this effect for an \texttt{IO} computation context will
execute the required primitive I/O actions, as well as throwing an exception
if any operation fails.

The following program type checks, and therefore implicitly carries 
a proof that the file resource protocol is followed correctly:

\begin{Verbatim}
testFile : Eff IO [FILE_IO (), STDIO] () 
testFile = catch (do open "testFile" Read
                     str <- readLine
                     close
                     putStrLn str)
                 (\err => putStrLn ("Error: " ++ show err))
\end{Verbatim}

The type of \texttt{testFile} states
that File I/O and console I/O are available effects, and in particular that
the resource associated with the File I/O will be in the same state on entry
and exit.
Therefore, attempting to write to the file, or failing to open or close the
file, would cause a \emph{compile-time} error.

We will use this technique extensively
throughout this paper: describe a resource usage protocol in terms of
state transitions; implement an effect which captures that protocol; implement
programs which, by using this effect, implicitly carry a proof that the resource
protocol has been correctly followed.

% =================================================
% =================================================

\section{Modelling resource usage protocols}
\label{rup}
In this section, we show how three such effects; CGI, database access and a simple session handler, may be implemented, and describe the benefits of developing programs using this technique over simply handling them in the IO context or as part of a monad transformer.

% =================================================

\subsection{CGI}
%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {Initialised};
  \node[state]         (C) [below of=B] {TaskRunning};
  \node[state]         (D) [below of=C] {TaskCompleted};
  \node[state]         (E) [below of=D] {HeadersWritten};
  \node[state]         (F) [below of=E] {ContentWritten};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{initialise}} (B)
        (B) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startTask}} (C)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{finishTask}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeHeaders}} (E)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{writeContent}} (F);
\end{tikzpicture}
}
\caption{CGI States}
\label{fig:cgistates}
\end{figure}

CGI is used to invoke an application on a web server, making use of environment variables to convey information gained from an HTTP request and using standard output to communicate with the remote client. Importantly, HTTP headers must be correctly written to the browser prior to any other output; failure to do so will result in an internal server error being shown.

A previous implementation of CGI in \idris{} implemented CGI an extension of monadic IO, as in Haskell. Whilst basic functionality worked correctly, this approach had several disadvantages; most notably, it was possible to perform arbitrary IO at any point in the program. If this were to happen, then the program would fail due to the fact that headers had not been written to the client.

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().
\begin{Verbatim}[samepage]
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure~\ref{fig:cgistates} shows the states through which the CGI program
progresses, and Figure~\ref{fig:cgieffect} shows how this is represented
as a resource-dependent algebraic effect. Each operation performed in an effectful
program requires the resource to be of a certain type, and the completion of
the operation may alter the type or value of the resource.

Upon creation, the CGI application is uninitialised, meaning that environment variables have not been queried to populate the CGI state. The only operation
that can be performed in this state is initialisation: by calling
\texttt{initialise}, a CGIInfo record is populated, and the state transitions
to \texttt{Initialised}. The \texttt{Init} operation is defined as part of the
\texttt{Cgi} effect, and involves transitioning from the uninitialised state to
the initialised state.

\begin{figure}[h]
\begin{Verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{Verbatim}
\caption{CGI Effect}
\label{fig:cgieffect}
\end{figure}

Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

User code executes in the \texttt{TaskRunning} state. Several operations, such as querying the POST and GET variables, are available in this state, alongside functions to output data to the web page and append data to the response headers. It is important to note, that at this stage nothing is written to the page, with the \texttt{output} and \texttt{addHeader} functions instead modifying the CGIInfo record. This data may then be printed at the end of the program's execution, in accordance with the resource usage protocol.

After the user code has finished execution, control returns to the library code. At this point, the state transitions to TaskCompleted, and the headers are written.
Finally, the headers and content are written which completes the process. Since we parameterise the resource over a state, we may ensure that certain operations only happen in a particular prescribed order.

In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We may therefore define the following type synonym, which we may use in order to make use of the CGI framework: 
\begin{Verbatim}
CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
 Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
\end{Verbatim}
This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.

A simple ``Hello, world!'' program would be defined as follows:

\begin{Verbatim}
module Main
import Cgi

sayHello : CGIProg [] ()
sayHello = output "Hello, world!"

main : IO ()
main = runCGI [initCGIState] sayHello
\end{Verbatim}
% =================================================

\subsection{Database access with SQLite}

%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {ConnectionOpened};
  \node[state]         (C) [below of=B] {PreparedStatementOpen};
  \node[state]         (D) [below of=C] {PreparedStatementBinding};
  \node[state]         (E) [below of=D] {PreparedStatementBound};
  \node[state]         (F) [below of=E] {PreparedStatementExecuting};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
\end{tikzpicture}
}
\caption{Database Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database engine often used as simple, structured storage for larger applications. We make use of SQLite due to its simplicity, although we envisage that these concepts would be applicable to more complex database management systems.

The creation, preparation and execution of SQL queries has a specific usage protocol, with several possible points of failure. Failure is handled in traditional web applications by the generation of exceptions, which may be handled in the program.
Handling such exceptions is often optional, however, and in some cases unhandled errors may cause a deployed web application to display an error to the user. Such errors can be used to determine the structure of an insecure SQL query, and are often used by attackers to determine attack vectors for SQL injection attacks.

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. This is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity).

\begin{figure}[h]
\begin{Verbatim}[samepage]
data Sqlite : Effect where
  OpenDB : 
    String -> 
    Sqlite () 
           (SQLiteRes ConnectionOpened) Bool
  CloseDB : 
    Sqlite (SQLiteRes ConnectionOpened) () Bool
  PrepareStatement : 
    String -> 
    Sqlite (SQLiteRes ConnectionOpened) 
           (SQLiteRes PreparedStatementOpen) Bool
  StartBind : 
    Sqlite (SQLiteRes PreparedStatementOpen) 
           (SQLiteRes PreparedStatementBinding) ()
  BindText : 
    ArgPos -> String -> Int -> 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBinding) 
           Bool
  FinishBind : 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBound) 
           Bool
  ExecuteStatement : 
    Sqlite (SQLiteRes PreparedStatementBound) 
           (SQLiteRes PreparedStatementExecuting) ()
  Finalise : 
    Sqlite (SQLiteRes PreparedStatementExecuting) 
           (SQLiteRes ConnectionOpened) 
           Bool
  GetColumnText : 
    Int -> 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           String
  RowStep : 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           StepResult
\end{Verbatim}
\caption{Database Effect}
\label{fig:dbeffect}
\end{figure}

There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
\begin{Verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

data DBPointer = ValidConn Ptr
               | InvalidConn

data StmtPointer = ValidStmt Ptr
                 | InvalidStmt 
\end{Verbatim}
If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}. At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety. The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}. In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function. Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function. We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor. 
\begin{Verbatim}
data StepResult = Unstarted
                | StepFail
                | StepComplete
                | NoMoreRows
\end{Verbatim}
Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource. Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore statically guarantee that only calls that will return a valid result are executed.

By incorporating pointers to open connections and prepared statements into the resource associated with the effect, we introduce a further layer of abstraction, in order to hide implementation details from the developer and encourage cleaner, less error-prone code.

\subsubsection{Example Code}
Programs making use of the DSL should look familiar to developers even without a background in functional programming. To demonstrate the functionality of the library, we return to the previous example of selecting the names and addresses of all staff working in a given department. Due to the fact that the \texttt{Effects} library overloads the bind operator, we may make use of \texttt{do}-notation, facilitating the usage of an imperative style.

We define a function of type:
\begin{Verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{Verbatim}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
\begin{Verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err 
\end{Verbatim}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

A call to \texttt{prepareStatement} attempts to create a prepared statement, and a subsequent call to \texttt{beginExecution} allows data to be retrieved from the database.

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
\begin{Verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] 
\end{Verbatim}
In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
\begin{Verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{Verbatim}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
\begin{Verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{Verbatim}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
\begin{Verbatim}
executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
                 (List DBVal)) -> 
  Eff m [SQLITE ()] (Either String ResultSet)
\end{Verbatim}
% =================================================
\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

A major strength of the \texttt{Effects} library is that it allows for simple composition of individual, fine-grained effects. By combining the individual CGI and SQLite components, we may construct a simple session handler to provide a notion of state across separate web requests.

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels

\begin{figure}[h]
\begin{Verbatim}[samepage]
data Session : Effect where
  LoadSession : 
  SessionID -> 
  Session (SessionRes SessionUninitialised) 
          (SessionRes SessionInitialised) 
          (Maybe SessionData)
  UpdateSession : 
    SessionData -> 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) ()
  CreateSession : 
    SessionData -> 
    Session (SessionRes SessionUninitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  DeleteSession : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool 
  WriteToDB : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool
  DiscardSessionChanges : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) ()
  GetSessionID : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  GetSessionData : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionData)
\end{Verbatim}
\caption{Session Effect}
\label{fig:sessioneffect}
\end{figure}

\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

  \tikzstyle{stateEdgePortion} = [black,thick];
  \tikzstyle{stateEdge} = [stateEdgePortion,->];
  \tikzstyle{edgeLabel} = [pos=0.5, font={\sffamily\small}];

  \node[initial,state] (A)              {SessionUninitialised};
  \node[state]         (B) [below of=A] {SessionInitialised};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}

Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated
with the session handler, and Figure \ref{fig:sessioneffect} the corresponding
algebraic effect. In this application, there exist two states:
In \texttt{SessionUninitialised}, the user may load or create a
session.
In \texttt{SessionInitialised}, the user may update the
representation of the session in memory, serialise the session and write it to
the database, or delete the session and invalidate the user's session key. The
introduction of these two states ensures that changes are explicitly either
written or discarded, eliminating the possibility of a developer updating the
session but neglecting to commit it to persistent storage. This, of course, is
under the assumption that the process exits cleanly: we attempt to facilitate
this by writing total functions where possible.

Much like the SQLite effect, we encapsulate failure by reflecting it in the
resource associated with the effect.

\begin{Verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

>>>>

@@

\begin{SaveVerbatim}{conn}

  $conn = new mysqli("localhost", "username", 
                     "password", "db");

\end{SaveVerbatim}
\useverb{conn}

\noindent
We then check to see if the connection was successful, and exit
if not.  This check is optional, so
it would be possible to omit it. However, this would cause
problems with later steps.

\begin{SaveVerbatim}{connerr}

  if (mysqli_connect_errno()) {
      exit();
  }

\end{SaveVerbatim}
\useverb{connerr}

\noindent
We then create a prepared statement detailing our query, and bind the
\texttt{`dept'} value:

\begin{SaveVerbatim}{prepare}

  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);

\end{SaveVerbatim}
\useverb{prepare}

\noindent
After the parameters have been bound, we execute the statement, assign
variables into which results will be stored, and fetch each row in turn. 
Failure to execute a statement before attempting to fetch rows would cause an error, as would attempting to execute a statement without binding variables to it.

\begin{SaveVerbatim}{stmtexec}

$stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }

\end{SaveVerbatim}
\useverb{stmtexec}

\noindent
Finally, once the statement and connection are no longer needed, they should be
closed in order to discard the associated resources:

\begin{SaveVerbatim}{connclose}

  $stmt->close();
  $conn->close();

\end{SaveVerbatim}
\useverb{connclose}

\noindent
Even in this small example, there exists a specific resource usage protocol.
Firstly, a connection to the database must be opened. The object-oriented style
used in the example encapsulates this to an extent, as the object must be
created in order for operations to be performed, however it is less obvious in
a procedural version of the code. Secondly, a prepared statement is created,
using the raw SQL and placeholders to which variables are later bound. The
statement is then executed, and each row is retrieved from the database.
Finally, the resources are freed.

@@

\idris{} \cite{brady2013idris} is a language with full dependent types, and
extensive support for EDSLs through overloading and syntax macros. Through the
use of \idris{}, and a framework for describing resource protocols using
\emph{algebraic effects}~\cite{brady:effects}, we
present a dependently-typed web framework, which allows the construction of
programs with additional guarantees about correctness and security, whilst
minimising the increase in development complexity.

@@

The code used to implement the framework and all associated examples used in this paper is available online at \url{http://www.github.com/idris-hackers/IdrisWeb}.
% =================================================
% =================================================
